---
title: "274lab06"
author: "Bochen Wang"
date: "2022-11-08"
output: html_document
---

(a)
Step 1 Data processing (make the data stationary)
y.12 <- diff(wine.bc, 12)

Step 2 Model identification
# Plot it:
par(mfrow=c(1, 1))
ts.plot(y.12, main="De-seasonlized Time Series",
ylab=expression(nabla[12]~Y[t]))
abline(h=mean(y.12), lty=2)

and use acf and pacf to identify
par(mfrow=c(1, 2))
acf(y.12, lag.max=60, main=expression(nabla[12]~Y[t]))
pacf(y.12, lag.max=60, main=expression(nabla[12]~Y[t])


Step 3 Model estimation
Step 4 Model selection
choose candidate models then Compute AICc 
then buillld the final model with fit

Step 5 Model diagnostics
# Residual plots:
res <- residuals(fit)
mean(res); var(res)
# layout(matrix(c(1, 1, 2, 3), 2, 2, byrow=T))

Step 6 Data forecast
use predict or use sarima.for() to forcast it.

(b)

```{r}
dowj_data <- scan("D:/RStudio/dowj.txt")
dowj <- ts(dowj_data)
# Plot data
ts.plot(dowj,main = "Dow Jones Index")
```
```{r}
dowj.diff <- diff(dowj,1)
ts.plot(dowj.diff, main = "De-trended data")

#check its residuals 
fit = arima(dowj, order=c(1,1,0), method="ML")
res <- residuals(fit)
hist(residuals(fit),main = "Histogram")
# q-q plot
qqnorm(residuals(fit))
qqline(residuals(fit),col ="blue")
```
the data seems detrended becasue its innovations follows a WN process and it seems normal

