---
title: "274hw05"
author: "Bochen Wang"
date: "2022-11-01"
output:
  pdf_document: default
  html_document: default
---

Problem 1
- define working directory;
setwd() and getwd()

– read and plot data;
read.csv() ts.plot()

– simulate and plot ARMA models;
x1 <- arima.sim(n = 100,model = list(ma=c(theta_1,theta_2))) plot(X1) ### add trend line and mean
  

– add trend and mean line to the original time series plot;
abline(lm(data~as.numeric(1:length(data))))
abline(h=mean(data)) 

ARMA(2,1)
# Simulate model
set.seed(4)
arma21 <- arima.sim(model = list(ar = c(-0.75,-0.5625),ma = 1.25,sd = 1),n = 200)
# Theoretical ACF
theo_acf <- ARMAacf(ar = c(-0.75,-0.5625), ma = 1.25,lag.max = 20, pacf = FALSE)
# Plot
op <- par(mfrow = c(1,2))
plot(theo_acf,type = "h",ylim = c(-1,1),
main = "Theoretical ACF",
col = "red",
ylab = expression(rho[X](h)),
xlab = "h")
abline(h = 0) # Add horizontal line
# Sample ACF
acf(arma21,lag.max = 20,
main = "Sample ACF",
ylim = c(-1,1),
xlab = "h",
ylab = expression(hat(rho)[X](h)))

– calculate and plot theoretical acf/pacf for ARMA models;
theo_acf <- ARMAacf(ar = c(-0.75,-0.5625), ma = 1.25,lag.max = 20, pacf = FALSE)


– calculate and plot sample acf/pacf;
acf(x, lag.max = NULL,
type = c(“correlation”, “covariance”, “partial”),
plot = TRUE, na.action = na.fail, demean = TRUE, . . . )
pacf(x, lag.max, plot, na.action, . . . )

acf(arma21,lag.max = 20,
main = "Sample ACF",
ylim = c(-1,1),
xlab = "h",
ylab = expression(hat(rho)[X](h)))

– check whether a particular model is causal/invertible (R commands to find and plot roots of polynomials)
polyroot(c()) ## see if they have unit roots
 
– perform Box-Cox transforms;
#for the dataset wine from lab 4
library(MASS)
t = 1:length(wine)
fit = lm(wine ~ t)
bcTransform = boxcox(wine ~ t,plotit = TRUE)

– perform differencing data at lags 1 and 12;
diff(x,lag = 1) diff(x, lag = 12)

– perform Yule-Walker estimation and find standard deviations of the estimates.
arima(data, order = c(p, 0, q), method = c("yule-walker"))

– perform MLE and check AICC associated with the model.
x_fit40 = arima(x, order = c(4,0,0), method = "ML")
library(qpcR)
AICc(x_fit40) 
#from lecture


Problem 2:

a:
I choose to analyze the Tesla company's stock prices from 2019 to today.
b:
I want to see if the electric car market have a future for entering this field.
c:
```{r}
tesla.csv <- read.csv("TESLA.csv")
tesla = ts(tesla.csv$Close, start= c(2019,1), frequency = 365)
plot.ts(tesla)
#it seems like a linear trend that is going up, in the middle it suddenly grows.

```
d:
```{r}

tesla_d <- diff(tesla.csv$Close, 3)
tesla_ts <- ts(tesla_d)
plot.ts(tesla_ts)
```
e:
```{r}
acf(tesla_ts,lag.max = 60,main = "")
pacf(tesla_ts,lag.max = 60,main = "")

```
here this data seem more like MA(2) model.

3:
```{r}
library(forecast)
x <- c(3.53,1.33,1.85,0.61,0.98,3.61,3.44,3.38,2.91,2.12,4.62,2.93)
ts_new <- ts(x, frequency = 4)
ndiffs(ts_new)
```

so we can take the values inside for all quarter 1:
$$
y_t = 2.637+0.252x_1+0.061x_2-0.202x_3\\
=2.637+0.252(3.53)+0.061(0.98)-0.202(2.91)\\
=2.637+0.88956+0.05928-0.58782
=2.9985
$$
so we choose A

Problem 4:
Becasue it is AR with mean 0
for $$ρ(2) = 0.215$$ and $$ρ(3) = −0.100$$
, so $$φ = −p(0.215)$$, also ρ(3) is negative.

$$X_{T+1} = φX_T$$ is -0.4637*-0431=0.19985

Problem 5:
$$
AIC_{AR(1)} = -2(-650)+2(1+2)=1306\\
AIC_{ARMA(1,1)} = -2 (-641)+2(1+1+2)=1290\\
AIC_{ARMA(1,2)} = -2(-636)+2(1+2+2)= 1282\\
AIC_{ARMA(2,3)} = -2 (-630)+2(2+3+2)= 1274\\
AIC_{ARMA(4,3)} = -2*(-629)+2(4+3+2)= 1276
$$
by the scores we can see ARMA(2,3) have the lowest thus the best model

GE 1:





















..